TBA

### 2020-08-24 manuscript.tex
Updates 0823

%\usepackage[draft]{todonotes}   % notes shown
\usepackage[disable]{todonotes} % notes hidden

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology} (cite more). Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales \citep{MhatreV.HoJi-AnnLee2012NIHAccess}, and dynamic DNA nanotechnology has been used to perform basic calculation \citep{Bui2018} and to probe single molecules via temporal DNA signatures \citep{Shah2019}. Both technologies rely on the hybridization reaction between complementary (or similar) DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions. Although many experimental and computational studies have been performed to investigate hybridization and dissociation, the sequence-dependent mechanisms of hybridization dynamics are not fully understood (cite). Moreover, it is unclear whether these processes occur in an "all-or-nothing" fashion or if some metastable states facilitate the transition. The stability of out of register or "slipped" base pairing in repetitive sequences has been well documented in previous computational studies \citep{Phys2014}, \citep{Xiao2019}) and are referred to as slip-strand DNA for longer seqeunces in vivo (cite, maybe too unrelated). Sandstead et al suggested the existence of frayed metastable states during duplex dissociation, where the stability of these states was dictated by oligonucleotide sequence \citep{Sanstead2016}. In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to  metastable structures mentioned above.

Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, \citep{Hoefert2011MolecularOligonucleotides}  and forward flux sampling  \citep{Phys2014} (cite more). Other studies use dramatically elevated or lowered temperatures to induce one way dissociation and hybridization events, respectively, \citep{Wong2008TheSimulations} (cite more). Experimental studies often employ temperature jump or other perturtabitive methods to drive DNA our of equilibrium and monitor relaxation processes (cite early T-jump and others). Single molecule analysis facilitate equilibrium analysis, however these present technical difficulties and are hampered by long timescales. In this work, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk}--  to run many unbiased coarse-grained simulations and combine these independent simulations to form an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstates distribution of DNA hybridization  \citep{Jin2019} \citep{Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. Pinamonti et al. used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

Our analysis reflects similar results to previous computational and experimental DNA work, while elucidating some new insights into sequences dependent dynamics and relative timescales. (something here about how this could motivate future experimental work or applications?) By employing SRVs to generate an optimized low dimensional basis, we can access higher resolution MSMs (shorter lag time) and generate a more detailed model. The slow dynamical modes learned by SRVs can be directly compared between sequences to determine otherwise subtle trends. More summarizing results here .....

@@

\subsection{\label{sec:methods}3spn2 Model}
Not sure if this is necessary, but can include a quick summary of model, why we chose to use it, and interesting features.
-Compatible with lammps
-Enough resolution to deserve dynamical behavior, coarse enough to run equlibrium sims
-Previous hybridization experiments show good agreement with experiments

@@

We initialized four sequences previously investigated by Sandstead et al \citep{Sanstead2016} -- AT-all, GC-core, GC-end, and GC-mix -- (small figure here) along with their complementary strands according a Beta helices according to 3SPN.2 documentation \citep{Phys2014}. We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end to end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A). This translated to a box size of 77.74 A and an effective oligo concentration of 7 mM. We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions. We maintained a Debye-Huckel screening potential with an ionic strength of 240 mM -- corresponding to NaCl concentration -- to account for phosphate backbone interactions. These electrostatics preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}. We run our simulations in the NVE ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. Simulation temperature was determined by sequence specific melting temperature such that strands were about equally likely to spend time in the hybridized or dissociated state. Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps. A 15 fs timestep was used for all simulations runs. For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation. An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state. In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}. We compared the VAMP-2 score (SI ?) for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set. Furthermore, we found an increase in VAMP-2 score when using reciprocal pairwise distances and chose these reciprocal permutation-free coordinates as a consistent feature set. These features were normalized and passed into sequence-specific SRVs.

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features. This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator can be estimated (equation here?). The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct markov state models. Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Harrigan2017LandmarkDynamics}. SRVs provide robust nonlinear approximation and computation timescales linearly with the amount of input data. This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein (cite). For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

Using optimized hyperparameters and featurized trajectory data, we generated a hierarchical dynamic encoder (HDE) to transform 55 reciprocal pairwise distances into a low dimensional SRV feature set. In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes. We determined the number of slow mode via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}. In particular we looked for a diminishing increase in the vamp-2 score, inconsistency between cross validation scores -- suggesting that the model maybe learning different modes, and a widening gap between train and test scores -- suggesting that the model might be overfitting on statistical noise in the training data. We used a batch size of 50000 and ran each model for total of 30 training epochs. We used two hidden layers and set the size of each layer to 100. For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training. SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.

@@

\subsection{\label{sec:Results}AT-all}
\subsubsection{\label{sec:Results}Explanation of state selection and slow modes}

In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes. For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step by step. Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}. We found that most implied timescales (with the exception of the leading timescale) converge at a lag time of 600 ns. We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation (*not sure if this warrants further discussion). It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency. Next we selected an optimal number of SRV components to include in our analysis. After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}. It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}. For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest. For the AT-all sequence, we generate SRVs with five components, representing the five slowest processes in the systems. Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands. The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by each process. For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal). Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the dynamics of the overall hybridization and dissociation process. The next four SRV components all show a relatively high correlation along different offset diagonals. These diagonals correspond to the intermolecular distances between shifted base pairs (SI figure for this?), and point to the existence of gradually faster sliding or "slithering" processes. Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than some  underlying fast dynamics \citep{Markegard2015}, \citep{Xiao2019}.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_srv_correlations.png}
        \caption{Shows Pearson correlations between leading SRV modes and all 100 intermolecular distances (needs labels for slow modes)}
        \label{fig:AT-all_srv_correlations}
	\end{center}
\end{figure}

From this analysis we can determine that the slowest dynamics are fully characterized by the hybridization process and slithering behavior of out of register base pairs. Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM. Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates. We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score. Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process. In the figure (dynamical figure) we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes. This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis. Using an MSM lag time of 1.2 ns, we then built a Bayesian MSM to calculate transition probability matrix between each microstate. Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represent a collection of metastable structures. Previous work have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels}, \citep{PinamontiAnalyzedModels}. Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence. For AT-all, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs. To visualize these six macrostates, we project the data into the two leading TICA coordinates. Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}. It is clear in figures (thermo fig) that the PCCA state assignments are capturing free energy minima in TICA space as independent states. After assigning these macrostates we can then calculate their relative probabilities and free energies, visualize representative molecular renderings, and estimates transition probabilities between states. In this final step, we use a minimum flux cutoff of 2e-6 in order to mitigate erroneous quick transition or skipping between states.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_dynamic.png}
        \caption{SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns}
        \label{fig:AT-all_dynamic}
	\end{center}
\end{figure}

In our coarse-grained SRV-MSM, we observe an "aligned hybridized" state, dissociated state, and four "shifted" states characterized by different combinations of out-of-register base pairings. These shifted states consist of 2 or 4 base pair shifting (single or doubled shifted) in either the 5' or 3' direction with varied stability and transition fluxes between kinetically neighboring states. This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019}, \citep{Phys2014}. We found a substantial difference in thermodynamic stability and kinetic behavior between the 5'A shifted states and 3'T shifted states. This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length. It has been shown that two base pair 5' dangling ends (or say one base pair inert tails) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes}, \citep{Dickman2012ThermodynamicDNAs}. This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends. Furthermore, Santalucia thermodynamic calculations reveal that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \citep{Allawi1998NearestDNA} (NN calculations SI?). Taken together, these experimental results substantiate why we observe the single shifted 5' state to be only slightly less stable relative to the aligned hybridized state. Moreover, it is expected that implied timescales corresponding to this process is on the same order of magnitude at the hybridization timescale. The stability of the single shifted 3' state is notably lower and on the same order of magnitude as the double shifted 5' state despite it having two less intact base pairs. The final state is more stable than one might expect given the difference in stability between the single and double 5' shifted states. Experimental work  \citep{Doktycz1990ThermodynamicATGC} studied four base pair dangling ends on DNA hairpins, and found that TATA ends are most stabilizing than ATAT. Although NN interactions paly a more important role in determining stabiity, this could explain why the relative free energy difference between single shifted 3' state and double shifted 3' is not as large as the difference between the single shifted 5' and double shifted 5' states.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_thermo.png}
        \caption{SRV-MSM TICA projection, state free energies, and transitions probabilities}
        \label{fig:AT-all_thermo}
	\end{center}
\end{figure}

>>>>

%\usepackage[draft]{todonotes}   % notes shown
\usepackage[disable]{todonotes} % notes hidden
\usepackage{comment}

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems.pdf, Rothemund2006FoldingPatterns, Gu2010ALine}. Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}. Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions. Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition. Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates. The stability of certain intermediates states, such as out-of-register or "slipped" base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014}, \citep{Xiao2019}). Furthermore, Sandstead et al suggested the existence of frayed metastable states during duplex dissociation, where the stability of these states was dictated by G:C bond placement in the 10-mer oligonucleotide sequence \citep{Sanstead2016}. In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to  metastable structures mentioned above.

Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Other computational studies use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Experimental studies often employ temperature jump, concentration jumps, or other perturbative methods to drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA}. Single molecule diffusion and tethered multifluorophore assays facilitate equilibrium analysis, however these present technical difficulties and are hampered by long timescales and data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices}. Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}.

In this work, we use the coarse-grain 3 site per nucleotide (3spn2) model to simulate hybridization and dissociation behavior at each sequence's melting temperature \citep{Hinckley2013AnHybridization}.  We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of meta-stable states during the hybridization and dissociation process. Furthermore we can perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (shorter lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs. Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequences dependent dynamics, meta-stable structures, and relative timescales.

@@

\subsection{\label{sec:methods}3spn2 Model}

Is some additional information warranted on the model itself before going into this specific setup?

@@

We initialized four sequences previously investigated by Sandstead et al. -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}. We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A). This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM. We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions. We maintained a Debye-Huckel screening potential with an ionic strength of 240 mM -- corresponding to NaCl concentration -- to account for phosphate backbone interactions. These electrostatics preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}. We run our simulations in the NVE ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. Simulation temperature was determined by sequence specific melting temperature such that strands were equally likely to spend time in the hybridized or dissociated state. Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps. A 15 fs timestep was used for all simulations runs. For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation. An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state. In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}. We compared the VAMP-2 score for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set. Furthermore, we found an increase in VAMP-2 score when using reciprocal pairwise distances and chose these reciprocal permutation-free coordinates as a consistent feature set. These features were normalized and passed into sequence-specific SRVs.

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features. This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator can be estimated (equation here?). The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs. Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Harrigan2017LandmarkDynamics}. SRVs provide robust nonlinear approximation and computation timescales linearly with the amount of input data. This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen}, \citep{Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

Using optimized hyperparameters and featurized trajectory data, we generated a hierarchical dynamic encoder (HDE) to transform 55 reciprocal pairwise distances into a low dimensional SRV feature set. In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes. We determined the number of slow mode via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}. In particular we looked for a diminishing increase in the vamp-2 score, inconsistency between cross validation scores -- suggesting that the model maybe learning different modes, and a widening gap between train and test scores -- suggesting that the model might be overfitting on statistical noise in the training data. We used a batch size of 50000 and ran each model for total of 30 training epochs. We used two hidden layers and set the size of each layer to 100. For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training. SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs. SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com} \citep{Abadi2016TensorFlow:Systems}.

@@

\subsection{\label{sec:Results}AT-all}

\subsubsection{\label{sec:Results}SRV optimization and analysis}

In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes. For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step-by-step. Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}. We found that most implied timescales (with the exception of the leading timescale) converge at a lag time of 600 ns. We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation. It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency. Next we selected an optimal number of SRV components to include in our analysis. After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}. It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}. For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest. As shown in (Figure \ref{fig:srv_crossval}), we see a clear convergence of the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis. Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands. The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by (or effective on) each process. For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal). Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the dynamics of the overall hybridization and dissociation process. The next four SRV components all show a relatively high correlation along different offset diagonals. These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes. Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics \citep{Markegard2015}, \citep{Xiao2019}.

\subsubsection{\label{sec:Results}SRV-MSM construction and optimization}

From this analysis we can determine that the slowest dynamics are fully characterized by the hybridization process and slithering behavior of out of register base pairs. Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM. Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates. We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score. Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process. In Figure \ref{fig:AT-all_dynamic} we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes. This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis. Using an MSM lag time of 1.2 ns, we then built a Bayesian MSM to calculate transition probability matrix between each microstate. Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represent a collection of metastable structures. Previous work have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels}, \citep{PinamontiAnalyzedModels}. Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence. For AT-all, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs. To visualize these six macrostates, we project the data into the two leading TICA coordinates. Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}. It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states. After assigning these macrostates we can then calculate their relative probabilities and free energies, visualize representative molecular renderings, and estimates transition probabilities between states. In this final step, we use a minimum flux cutoff of 2e-6 in order to mitigate erroneous quick transition or skipping between states.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_dynamic.png}
        \caption{AT-all SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns. Leading slow mode correlations to all 100 intermolecular distances between oligos.}
        \label{fig:AT-all_dynamic}
	\end{center}
\end{figure}

\subsubsection{\label{sec:Results}Implied timescales for TICA-MSMs for }

In our coarse-grained SRV-MSM, we observe an "aligned hybridized" state, dissociated state, and four "shifted" states characterized by different combinations of out-of-register base pairings. These shifted states consist of 2 or 4 base pair shifting (single or doubled shifted) in either the 5' or 3' direction with varied stability and transition fluxes between kinetically neighboring states. This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019}, \citep{Phys2014}. We found a substantial difference in thermodynamic stability and kinetic behavior between the 5'A shifted states and 3'T shifted states. This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length. It has been shown that two base pair 5' dangling ends (or dangling ends with one inert tail) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes}, \citep{Dickman2012ThermodynamicDNAs}. This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends. Furthermore, Santalucia thermodynamic calculations (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \citep{Allawi1998NearestDNA} (Note that these calculations are performed at higher salt concentration, and that salt seems to have a significant effect on the impact of inert tails on stability). Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state. Furthermore, we expect implied timescales of the two leading modes to be on the same order of magnitude as these reflect the average lifetimes of these states. The stability of the single-shifted 3' state is notably lower than that of the single-shifted 5' state. In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having more less intact base pairs. This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below. The final state is more stable than one might expect given the difference in stability between the single and double 5' shifted states. This might be attributed to the asymptotic effects of inert tail destabilitization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}. Accordingly, the longer inert tails in the double-shifted states have less of an effect on relative 5' vs. 3' stability compared to the shorter tails in the single-shifted states.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.PNG}
        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
        \label{fig:AT-all_in_one}
	\end{center}
\end{figure}

\subsubsection{\label{sec:Results}Transition probabilities between states}
